---
title: "Data Cleaning"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    theme: cosmo
execute:
  warning: false
  message: false
---
```{r setup, include=FALSE}
library(tidyverse)
library(skimr)
library(naniar)

miami_raw <- read_csv("data/raw/miami-housing.csv")
```

## Raw Data Overview
```{r raw-overview}
glimpse(miami_raw)
```

The raw dataset contains `r nrow(miami_raw)` properties and `r ncol(miami_raw)` columns. Each row represents a single residential property sold in Miami in 2016.

## Column Name Standardization
```{r clean-names}
library(janitor)

miami <- miami_raw %>%
  clean_names() %>%
  rename(plane_noise = avno60plus) %>%
  select(-parcelno)

names(miami)
```

`parcelno` is a unique property identifier with no predictive value and is dropped. `avno60plus` is renamed to `plane_noise` for readability.

## Type Corrections
```{r type-corrections}
miami <- miami %>%
  mutate(
    plane_noise       = factor(plane_noise),
    month_sold        = factor(month_sold),
    structure_quality = factor(structure_quality)
  )

miami %>%
  summarise(across(everything(), class)) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "type")
```

Three variables are stored as integers in the raw data but are categorical in nature. Converting them to factors ensures they are dummy-encoded correctly during preprocessing rather than treated as continuous.

## Missingness Check
```{r missingness}
miss_var_summary(miami) %>%
  filter(n_miss > 0)
```
```{r missingness-confirm}
cat("Total missing values:", sum(is.na(miami)))
```

No missing values are present. No imputation is needed.

## Outcome Variable Distribution
```{r outcome-dist}
p1 <- ggplot(miami, aes(x = sale_prc)) +
  geom_histogram(bins = 60, fill = "steelblue", color = "white") +
  scale_x_continuous(labels = scales::dollar_format()) +
  labs(title = "Original scale", x = "Sale price (USD)", y = "Count")

p2 <- ggplot(miami, aes(x = log(sale_prc))) +
  geom_histogram(bins = 60, fill = "steelblue", color = "white") +
  labs(title = "Log scale", x = "log(Sale price)", y = "Count")

library(patchwork)
p1 + p2 +
  plot_annotation(title = "Sale price before and after log transformation")
```

Sale price is heavily right-skewed with a long tail of high-value properties. A log transformation produces an approximately normal distribution, which improves model performance and makes residuals easier to interpret.

## Final Cleaned Data
```{r final-clean}
miami <- miami %>%
  mutate(sale_prc_log = log(sale_prc)) %>%
  select(-sale_prc)

skim(miami)
```
```{r save}
write_rds(miami, "data/processed/miami_clean.rds")
cat("✓ miami_clean.rds written to data/processed/\n")
```

The cleaned dataset contains `r nrow(miami)` rows and `r ncol(miami)` columns, with 3 factor variables, 12 numeric predictors, and the log-transformed outcome variable.