---
title: "Exploratory Data Analysis"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    theme: cosmo
execute:
  warning: false
  message: false
---
```{r setup, include=FALSE}
library(tidyverse)
library(patchwork)

miami <- read_rds("data/processed/miami_clean.rds")
```

## Outcome Variable
```{r outcome}
p1 <- ggplot(miami, aes(x = exp(sale_prc_log))) +
  geom_histogram(bins = 60, fill = "steelblue", color = "white") +
  scale_x_continuous(labels = scales::dollar_format()) +
  labs(title = "Original scale", x = "Sale price", y = "Count")

p2 <- ggplot(miami, aes(x = sale_prc_log)) +
  geom_histogram(bins = 60, fill = "steelblue", color = "white") +
  labs(title = "Log scale", x = "log(Sale price)", y = "Count")

p1 + p2 +
  plot_annotation(title = "Sale price distribution before and after log transformation")
```

## Location Effects
```{r location}
p1 <- ggplot(miami, aes(x = latitude, y = sale_prc_log)) +
  geom_point(alpha = 0.1, size = 0.5) +
  geom_smooth(color = "red") +
  labs(title = "Latitude vs log price", x = "Latitude", y = "log(Sale price)")

p2 <- ggplot(miami, aes(x = longitude, y = sale_prc_log)) +
  geom_point(alpha = 0.1, size = 0.5) +
  geom_smooth(color = "red") +
  labs(title = "Longitude vs log price", x = "Longitude", y = "log(Sale price)")

p1 + p2
```

The latitude relationship is clearly nonlinear — prices peak in the middle latitudes and drop off at both ends. The longitude relationship shows a U-shape, motivating the `longitude^2` term in the EDA recipe.

## Distance Variables
```{r distances}
p1 <- ggplot(miami, aes(x = ocean_dist, y = sale_prc_log)) +
  geom_point(alpha = 0.1, size = 0.5) +
  geom_smooth(color = "red") +
  labs(title = "Ocean distance", x = "Distance (ft)", y = "log(Sale price)")

p2 <- ggplot(miami, aes(x = log(ocean_dist), y = sale_prc_log)) +
  geom_point(alpha = 0.1, size = 0.5) +
  geom_smooth(color = "red") +
  labs(title = "log(Ocean distance)", x = "log(Distance)", y = "log(Sale price)")

p3 <- ggplot(miami, aes(x = hwy_dist, y = sale_prc_log)) +
  geom_point(alpha = 0.1, size = 0.5) +
  geom_smooth(color = "red") +
  labs(title = "Highway distance", x = "Distance (ft)", y = "log(Sale price)")

p4 <- ggplot(miami, aes(x = log(hwy_dist + 1), y = sale_prc_log)) +
  geom_point(alpha = 0.1, size = 0.5) +
  geom_smooth(color = "red") +
  labs(title = "log(Highway distance)", x = "log(Distance)", y = "log(Sale price)")

(p1 + p2) / (p3 + p4)
```

Log-transforming highway distance straightens the relationship considerably. Ocean distance remains curved even after logging, motivating the natural spline in the EDA recipe.

## Special Features Value
```{r spec-feat}
p1 <- ggplot(miami, aes(x = spec_feat_val, y = sale_prc_log)) +
  geom_point(alpha = 0.1, size = 0.5) +
  geom_smooth(color = "red") +
  labs(title = "Special features value", x = "Value ($)", y = "log(Sale price)")

p2 <- ggplot(miami, aes(x = sqrt(spec_feat_val), y = sale_prc_log)) +
  geom_point(alpha = 0.1, size = 0.5) +
  geom_smooth(color = "red") +
  labs(title = "sqrt(Special features value)", x = "sqrt(Value)", y = "log(Sale price)")

p1 + p2
```

The square root transformation reduces the right skew and produces a more linear relationship with log price.

## Structural Variables
```{r structural}
p1 <- ggplot(miami, aes(x = tot_lvg_area, y = sale_prc_log)) +
  geom_point(alpha = 0.1, size = 0.5) +
  geom_smooth(color = "red") +
  labs(title = "Living area (sqft)", x = "Sq footage", y = "log(Sale price)")

p2 <- ggplot(miami, aes(x = age, y = sale_prc_log)) +
  geom_point(alpha = 0.1, size = 0.5) +
  geom_smooth(color = "red") +
  labs(title = "Property age", x = "Age (years)", y = "log(Sale price)")

p3 <- ggplot(miami, aes(x = structure_quality, y = sale_prc_log)) +
  geom_boxplot(fill = "steelblue", alpha = 0.7) +
  labs(title = "Structure quality", x = "Quality rating", y = "log(Sale price)")

p4 <- ggplot(miami, aes(x = lnd_sqfoot, y = sale_prc_log)) +
  geom_point(alpha = 0.1, size = 0.5) +
  geom_smooth(color = "red") +
  labs(title = "Land area (sqft)", x = "Sq footage", y = "log(Sale price)")

(p1 + p2) / (p3 + p4)
```

## Correlation Overview
```{r correlation}
miami %>%
  select(where(is.numeric)) %>%
  cor() %>%
  as.data.frame() %>%
  rownames_to_column("var1") %>%
  pivot_longer(-var1, names_to = "var2", values_to = "correlation") %>%
  ggplot(aes(x = var1, y = var2, fill = correlation)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Correlation matrix — numeric predictors", x = NULL, y = NULL)
```